---
title: "Lab: NYC + Covid-19 + Census Data"
output: 
  learnr::tutorial:
  progressive: true
  allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include = FALSE}
if(!require(learnr)) {
  install.packages("learnr")
}
library(learnr)

if(!require(dplyr)) {
  install.packages("dplyr")
}
library(dplyr)

if(!require(sf)) {
  install.packages("sf")
}
library(sf)

if(!require(viridis)) {
  install.packages("openintro")
}
library(viridis)

if(!require(shiny)) {
  install.packages("shiny")
}
library(shiny)

if(!require(tidyr)) {
  install.packages("tidyr")
}
library(tidyr)

borough <- read.csv("data/data_summ_bor.csv")
zcta <- read.csv("data/data_summ_zcta.csv")
NYC_zips <- st_read("Shape_Files/ZIP_CODE_040114.shp")
NYC_zips <- NYC_zips$geometry

getShapeColors <- function(data, nbreaks=5, missingcol="gray88", rdgrbl=c(0,0,1), mycolors=NULL, max_opac=0.55){
  if(length(nbreaks) == 1) {
    mybreaks <- c(-1, quantile(data,
                               probs = seq(0, 1, length.out=nbreaks+1),
                               na.rm = T))
  } else {
    mybreaks <- c(-1,nbreaks)
  }
  if(!is.null(mycolors)){
    mycols <- mycolors
  } else {
    mycols <- rgb(rdgrbl[1],rdgrbl[2],rdgrbl[3],
                  seq(0.05,max_opac, length.out = length(mybreaks)-1))
  }
  mycols[1] <- missingcol
  data[is.na(data)] <- -.1
  mycolorscheme <- cut(data, mybreaks) %>%
    as.numeric()
  mycolorscheme <- mycols[mycolorscheme]
  levs <- levels(cut(data, mybreaks))
  levs[1] <- "No Data"
  return(list(mycolorscheme, levs, mycols))
}

zcta2 <- zcta[!duplicated(zcta$MODZCTA),]
propwhite_take2 <- zcta2 %>% 
  transmute(White_Prop = round(White_Total/Race_Total*100, 2),
         Cases_per_1000 = round(Positive/Total_Pop*1000, 2)) %>%
  na.omit() %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 2),
         CaseRate_Lev = ntile(Cases_per_1000, 2)) %>%
  select(White_Prop_Lev, CaseRate_Lev) %>%
  table()

thresholdValue <- propwhite_take2[2,1]

reducedZCTA <- zcta2 %>% 
  transmute(White_Prop = round(White_Total/Race_Total*100, 1), 
            Cases_per_1000 = round(Positive/Total_Pop*1000, 2)) %>%
  na.omit()
# 
# set.seed(333)
simvals <- vector()
# for(i in 1:1000){
#   # reshuffle the proportion white-
#   reducedZCTA$White_Prop <- sample(reducedZCTA$White_Prop)
#   # then recompute median split
#   simdata <- reducedZCTA %>% 
#     transmute(White_Prop_Lev = ntile(White_Prop, 2),
#               CaseRate_Lev = ntile(Cases_per_1000, 2)) 
#   simvals[i] <- table(simdata)[2,1]
# }

knitr::opts_chunk$set(echo = FALSE)
```  
  
```{r load-data}
zcta <- zcta[!duplicated(zcta$MODZCTA),]
```

```{r load-data2}

```
  
## The data: a summary 
In this lab, you will inspect and interact with some datasets that have been compiled for you. The datasets loaded here were compiled using American Community Survey (ACS) census data, NYC COVID19 data (downloaded on April 23, 2020 from here: https://github.com/nychealth/coronavirus-data), and shape files for NYC zip codes (downloaded from here: https://data.cityofnewyork.us/widgets/i8iw-xf4u). You can see how all of these data were created, including how to add your own variables or change the regions, by reading the accompanying tutorials in more detail. The purpose of this lab is to give you a chance to run some code and play with some of the data, while bypassing the need to download and load everything on your own computer. There are three files loaded for you in this space:

+ zcta - data by zip code (zipcode tabulation area),
+ borough - data by borough
+ NYC_zips - a shape file used only for plotting maps of nyc zip codes 


## Inspecting the data  

### Exercise: Print the first few rows

Use the `head()` function or square brackets to inspect the first 5 rows of the `borough` and `zcta` datasets. 
```{r inspect_data, exercise=TRUE, exercise.setup = 'load-data'}

```

```{r inspect_data-solution}
head(borough, 5)
head(zcta, 5)
#or:
borough[1:5,]
zcta[1:5,]
```
   
```{r 5row-quiz}
quiz(
  question("What is the median income in the Bronx?",
    answer("36381"),
    answer("59620"),
    answer("46586"),
    answer("27074", correct = TRUE)
  )
)

```  

### Exercise: Inspect the structure of the data    

Try using the `str()` function below to look at the structure of each dataset. The term "structure" refers to the organization of a data set, which can be more complicated than a simple table. For example, it could have variables nested in other variables. When the data are in a table, `str()` will tell you about the names and types of variables.

```{r structure, exercise=TRUE, exercise.setup = 'load-data'}

```

```{r structure-solution}
str(borough)
str(zcta)
```
     
At this point, it might be useful to also inspect some of the variables more carefully to make sure we understand what they are. For now, let's focus on the zipcode level data. Looking at the data structure, we note that the second three variables are called "Positive," "Total," and "zcta_cum.perc_pos". "Positive" gives the number of positive tests by zipcode, "Total" gives the number of total tests by zipcode. We can confirm below that zcta_cum.perc_pos is just the percentage of total tests that were positive (i.e., Positive/Total*100, rounded to two decimals):    
```{r postotal, exercise=TRUE, exercise.setup = 'load-data'}
head(round(zcta$Positive/zcta$Total*100,2))
head(zcta$zcta_cum.perc_pos)
```
  
Another useful consistency check: the census data includes a "total" variable for most demographic variables (i.e., sex, age, income, race, education). In this dataset, we have collected the reported values for "Race_Total" (i.e., total of all race categories). Does this show the same number as the total population from the SEX BY AGE tables? Perhaps estimates based on race do not count everybody? Upon further inspection, these variables actually do have the same values:   

```{r pop, exercise=TRUE, exercise.setup = 'load-data'}
head(zcta$Race_Total)
head(zcta$Total_Pop)
```
  
That said, you might notice that there is another variable called Total_EducPop. This variable was imported as part of the group of variables related to educational attainment level, including Bachelors, Masters, Professional, and Doctorate degrees. 

```{r pop2, exercise=TRUE, exercise.setup = 'load-data'}
head(zcta$Total_Pop)
head(zcta$Total_EducPop)
```
  
Note that the values in this column are smaller than the Population (equals Race_Total) columns. Looking at the documentation provided by the Census for this group of variables, it turns out the variables we were examining (college education) are recorded only for people over the age of 25.  So the moral of the story is, don't always trust the names of the variables created by someone else (in this case, us)!  

## Case rates by zip code 

This article from the intercept (https://theintercept.com/2020/04/09/nyc-coronavirus-deaths-race-economic-divide/) starts with two tables showing zip codes with the highest and lowest positive test rates. We can reproduce these tables using the code below (they look slightly different because our data is more recent than that used in the article and we are reporting median income rather than per-capita income). Nonetheless, they look fairly similar (and we see a similar pattern with respect to income disparities).
  
```{r pop4, exercise=TRUE, exercise.setup = 'load-data'}
#5 highest
zcta %>% 
  mutate(Percent_Positive = round(Positive/Total*100,2)) %>%
  select(MODZCTA, Total, Positive, Percent_Positive, Med_Income) %>%
  arrange(desc(Percent_Positive)) %>%
  head(5)

#5 lowest
zcta %>% 
  mutate(Percent_Positive = round(Positive/Total*100,2)) %>%
  select(MODZCTA, Total, Positive, Percent_Positive, Med_Income) %>%
  arrange(Percent_Positive) %>%
  head(5)
```   
   
The tables that we produced above examine the percent of *tests* that come back positive. But what if we look at positive tests as a percentage of the total *population* by zip code. Does it make a difference? See if you can alter the code above so that you can find the zip codes with the highest positive case rate (defined as the percentage of the total population that has tested positive).    
```{r pop5, exercise=TRUE, exercise.setup = 'load-data'}

```
    
    
```{r pop5-solution}
#5 highest
zcta %>% 
  mutate(Pos_Case_Rate = round(Positive/Total_Pop,4)) %>%
  select(MODZCTA, Total_Pop, Positive, Pos_Case_Rate, Med_Income) %>%
  arrange(desc(Pos_Case_Rate)) %>%
  head(5)

#5 lowest
zcta %>% 
  mutate(Pos_Case_Rate = round(Positive/Total_Pop,4)) %>%
  select(MODZCTA, Total_Pop, Positive, Pos_Case_Rate, Med_Income) %>%
  arrange(Pos_Case_Rate) %>%
  head(5)
```
     
Once you do this, you'll see that it does make a difference in which zip codes appear (although we still see a similar relationship between case rate and income). If you look at the data on a broader scale, the percent of tests that come back positive decreases as the number of tests administered increases. Therefore, we could imagine a situation where wealthy areas have more access to tests, leading to a lower positive test rate (relative to other areas) but a high rate of positive cases relative to the whole population (because more positive cases are found). 

To see this, we can plot both the positive test rate and the positive case rate as a function of the overall testing rate:   


```{r pop6, exercise=TRUE, exercise.setup = 'load-data'}
plot(zcta$Total/zcta$Total_Pop*100, 
     zcta$zcta_cum.perc_pos,
     xlab = "percent of population tested",
     ylab = "positive TEST rate",
     main = "")
```
  
```{r pop7, exercise=TRUE, exercise.setup = 'load-data'}
plot(zcta$Total/zcta$Total_Pop*100, 
     zcta$Positive/zcta$Total_Pop*100,
     xlab = "percent of population tested",
     ylab = "positive CASE rate",
     main = "")
```

It seems like, for zip codes where more than 3% of the population is being tested, the positive test rate does decrease as the testing rate increases; however, the lowest positive test rates are still occuring in the zip codes with the lowest testing rates. That's all we'll say about this here, but please feel free to investigate further! The main take-away for now is that your denominator matters.  
  

## Tables by race and income  

Race is a complex subject and, in fact, race has many categorizations in the ACS dataset. A simplifying step, if we are intersted in associations between race and COVID-19 outcomes, is to compare one category against all the others. You could choose any race (e.g., proportions of non-Hispanic Asian are in the column NH_Asian). But we will concentrate on total population white, since a lot of the conversation around racial disparities centers on the degree to which a white majority is associated with privilege. 

Suppose, then, that we want to look at *racial composition* in terms of proportion white, and in particular, we want to discretize this racial composition variable into three groups: low, medium, and high proportion white. This is the kind of thing that happens in a lot of media reports.

The code for this transformation looks a little complicated, but you should be able to follow the general idea. The `mutate` function is used to create new variables in an existing table (there are other ways to do this in R). The `ntile` function assigns each zipcode to one of our three groups based on rank-ordering (it is an implementation of cuts using **quantiles**), and the `group_by` and `summarize` functions agreggate the positive case rate over all of the zipcodes in the same group.

```{r oneway_race, exercise=TRUE, exercise.setup = 'load-data'}
propwhite <- zcta %>% 
  mutate(White_Prop = round(White_Total/Race_Total,2)) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 3)) %>%
  group_by(White_Prop_Lev) %>% 
  summarise(Cases_per_1000 = round(sum(Positive)/sum(Total_Pop)*1000, 2)) %>% 
  na.omit()
propwhite
```


```{r oneway_race-solution}
propwhite <- zcta %>% 
  mutate(White_Prop = round(White_Total/Race_Total,2)) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 2)) %>%
  group_by(White_Prop_Lev) %>%
  summarise(Cases_per_1000 = round(sum(Positive)/sum(Total_Pop)*1000, 2)) %>%
  na.omit()
propwhite
```  
  
Notice that case rates are higher in zipcodes with a relatively low proportion of white people, and lower in zipcodes with a relatively high proportion of white people). Activity: Run the provided code. Then see if you can change it so that there are only 2 categories for the White_Prop variable.    

### Income

We can similarly define a one-way table of zipcodes summarized by low income, medium income, and high income. (Notice that case rates are higher in zip codes with a relatively low median income, and lower in zip codes with a relatively high median income):  
```{r oneway_inc, exercise=TRUE, exercise.setup = 'load-data'}
inc <- zcta %>% 
  mutate(Med_Inc_Lev = ntile(Med_Income, 3)) %>%
  group_by(Med_Inc_Lev) %>%
  summarise(Cases_per_1000 = round(sum(Positive)/sum(Total_Pop)*1000, 2)) %>%
  na.omit()
inc

# inc2 <- zcta %>% 
#   mutate(Med_Inc_Lev = ntile(Med_Income, 3)) %>%
#   group_by(Med_Inc_Lev) %>%
#   summarise(Cases_per_1000 = round(mean(Total_Pop),2)) %>%
#   na.omit()
# inc2


```
  
In the above examples of one-way tables, we take a single variable, such as proportion-white or median-income, and discretize that variable into levels. We then group the zipcodes (which are our data cases) by level, and display this COVID case-rate aggregated over the zipcodes in each level.

Question: based on the code that we used, are we "weighing" each zipcode by its population, or are we weighing each zipcode equally?


### Two variables at time

Let's now try to reproduce a 2-way table, looking simultaneously at the variables of race and income. To make a two way table, of course we need to create the two groupings (proportion-white and median-income) as before. Now when we use the `group_by` function, we include both variables.

We notice something kind of interesting. Notice that, among zip-codes with the lowest proportion of white residents, case rates are higher in zip codes with higher incomes, white the reverse is true for zip codes with a medium or high proportion of white residents.      
```{r twoway, exercise=TRUE, exercise.setup = 'load-data'}
inc_propwhite <- zcta %>% 
  select(White_Total, Race_Total, Med_Income, Positive, Total_Pop) %>% 
  na.omit() %>%
  mutate(White_Prop = round(White_Total/Race_Total, 3)*100) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 3)) %>%
  mutate(Med_Inc_Lev = ntile(Med_Income, 3)) %>%
  group_by(White_Prop_Lev, Med_Inc_Lev) %>%
  summarise(Cases_per_1000 = round(sum(Positive)/sum(Total_Pop)*1000, 2)) %>%
  tidyr::spread(White_Prop_Lev, Cases_per_1000, fill=0)

inc_propwhite[,1] <- c("Low Income", "Med Income", "High Income")
colnames(inc_propwhite) <- c("inc/race","Low Prop White", "Med Prop White", "High Prop White")
inc_propwhite
```
  
  
Bonus: we could take a closer look at what's going on here by plotting the positive case rate against median income and color coding by the three categories that we created for proportion of white residents (you might notice that there is a much wider range of median incomes among the zip codes with a high proportion of white residents):  
```{r twoway2, exercise=TRUE, exercise.setup = 'load-data'}
dat <- zcta %>% 
  select(White_Total, Race_Total, Med_Income, Positive, Total_Pop) %>% 
  na.omit() %>%
  mutate(White_Prop = round(White_Total/Race_Total, 3)*100) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 3))

plot(dat$Med_Income, 
     dat$Positive/dat$Total_Pop*100, 
     col=dat$White_Prop_Lev+3,
     xlab = "Median Income",
     ylab = "Case Rate",
     pch = 16)
legend("topright",
       col = c(4,5,6),
       pch = 16,
       legend = c("Low Prop White",
                  "Med Prop White",
                  "High Prop White"))
```
  

### Slicing the same data differently
  
For a moment, let's come back to the one-way table of the proportion of the population that is white. The point of that tabulation was to explore whether there was a relationship between race composition and COVID-19 incidence. 

Here is another way we could have conducted that analysis. In this version, we make a two-way table of race and case rate. That is, we also discretize case rate as, for example, low and high. In the cells of this table, we will count how many of the roughly 200 zipcodes fit the joint categories.

The code below does the following: First, we divide the zip codes into two equally sized groups based on the proportion of the population that is white. This is also called a median-split. Half of the zip codes are classified as being high proportion white, and half are classified as low proportion white Then, we do the same thing by case rate. Half of the zip codes are classified as having a high case rate and half are classified as having a low case rate. Finally, we construct a 2-way table of these new categorical variables, where the values in the table are counts of zip codes. For example, there are 38 zip codes that are classified as having a low case rate AND a low proportion of white people.    

```{r oneway_race2, exercise=TRUE, exercise.setup = 'load-data'}

propwhite_take2 <- zcta %>% 
  transmute(White_Prop = round(White_Total/Race_Total*100, 2),
         Cases_per_1000 = round(Positive/Total_Pop*1000, 2)) %>%
  na.omit() %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 2),
         CaseRate_Lev = ntile(Cases_per_1000, 2)) %>%
  select(White_Prop_Lev, CaseRate_Lev) %>%
  table()
propwhite_take2
```
  
We observe something that may or may not be surprising: There are more zip codes with a high proportion of white people and low case rate or low proportion of white people and high case rate (as compared to the other two categories). This suggests that zip codes with a higher proportion of non-white residents tend to have higher case rates. Can we test whether these differences are meaningful?

Let's start with the value in the "low proportion white, high case rate category" (which is `r propwhite_take2[2,1]`). 


We'll create a fake dataset that looks like the observed data (it would produce the same table as above), but then we will randomly reassign the levels of "proportion white" 1000 times, each time storing the number of zip codes in the low-white/high-case category. Then we can plot a histogram of the simulated values (with a vertical line at our observed value of `r propwhite_take2[2,1]`). We can also calculate the proportion of simulated values greater than or equal to `r propwhite_take2[2,1]`. See if you can follow the code below:  

```{r sim, exercise=TRUE, exercise.setup = 'load-data'}
thresholdValue <- propwhite_take2[2,1]

reducedZCTA <- zcta %>% 
  transmute(White_Prop = round(White_Total/Race_Total*100, 1), 
            Cases_per_1000 = round(Positive/Total_Pop*1000, 2)) %>%
  na.omit()

set.seed(333)
simvals <- vector()
for(i in 1:1000){
  # reshuffle the proportion white
  reducedZCTA$White_Prop <- sample(reducedZCTA$White_Prop)
  # then recompute median split
  simdata <- reducedZCTA %>% 
    transmute(White_Prop_Lev = ntile(White_Prop, 2),
              CaseRate_Lev = ntile(Cases_per_1000, 2)) 
  simvals[i] <- table(simdata)[2,1]
}


hist(simvals, xlim=c(35,65), main="Zipcode race permutation results",
     xlab="No. of cases with low-white and high-case split")
abline(v=thresholdValue, lwd=2, col="blue")

#proportion of simulated values greater than threshold
sum(simvals>=thresholdValue)/1000
```
  
We note that `r 100*sum(simvals>=thresholdValue)/1000`% of our simulated values were greater than `r thresholdValue`. It's therefore very unlikely that we would have observed this large of a value by random chance (if the truth was that there was no relationship between the proportion of the population that is white and the case rate). However, this cannot tell us WHY this relationship exists (there are many theories that have been proposed).   


### A quick side note:   

Another thing to take a look at: you may have noticed that we split case rate and proportion white into only two equally sized categories above. Here is the code again (except the code to specify the row and column names was taken out for now; see how to replace it in the "solution"). See if you can change the code to split the zip codes into 3, or even 4 categories. Do you notice a different pattern?  (Hint: if you have more than 2 categories, but only two labels for your row and column names, you will get an error!)
```{r oneway_race3, exercise=TRUE, exercise.setup = 'load-data'}
propwhite2 <- zcta %>% 
  mutate(White_Prop = round(White_Total/Race_Total*100, 2)) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 2)) %>%
  na.omit() %>%
  mutate(Cases_per_1000 = round(Positive/Total_Pop*1000, 2)) %>%
  mutate(CaseRate_Lev = ntile(Cases_per_1000, 2)) %>%
  group_by(White_Prop_Lev, CaseRate_Lev) %>%
  summarise(N_Zips = n()) %>%
  spread(White_Prop_Lev, N_Zips, fill=0)

# label the row and column names
propwhite2[,1] <- c("Low Case Rate", "High Case Rate")
colnames(propwhite2) <- c("case/race", "Low Prop White", "High Prop White")

propwhite2

```  
  
```{r oneway_race3-solution}
propwhite2 <- zcta %>% 
  mutate(White_Prop = round(White_Total/Race_Total*100, 2)) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 3)) %>%
  na.omit() %>%
  mutate(Cases_per_1000 = round(Positive/Total_Pop*1000, 2)) %>%
  mutate(CaseRate_Lev = ntile(Cases_per_1000, 3)) %>%
  group_by(White_Prop_Lev, CaseRate_Lev) %>%
  summarise(N_Zips = n()) %>%
  spread(White_Prop_Lev, N_Zips, fill=0)

# label the row and column names: 3 levels
propwhite2[,1] <- c("Low Case Rate", "Medium Case Rate", "High Case Rate")
colnames(propwhite2) <- c("case/race", "Low Prop White", "Medium Prop White", "High Prop White")

propwhite2
```
  
   
## Mapping NYC    
In order to create a map of NYC, with zip codes colored by a particular variable, one first needs to download a shape file. A zipcode level shape file has been loaded here for you and is called NYC_zips. For those who are interested, this was loaded with the st_read() function from the sf package. Next, to make the plot useful, it is possible to determine the color of each zip code on the map based on values of a particular variable. We have written a function to do this and have copied it below so you can inspect it. The function is loaded for you to use in subsequent activities.  
```{r map1, exercise=TRUE, exercise.setup = 'load-data2'}
getShapeColors <- function(data, nbreaks=5, missingcol="gray88", rdgrbl=c(0,0,1), mycolors=NULL, max_opac=0.55){
  if(length(nbreaks) == 1) {
    mybreaks <- c(-1, quantile(data,
                               probs = seq(0, 1, length.out=nbreaks+1),
                               na.rm = T))
  } else {
    mybreaks <- c(-1,nbreaks)
  }
  if(!is.null(mycolors)){
    mycols <- mycolors
  } else {
    mycols <- rgb(rdgrbl[1],rdgrbl[2],rdgrbl[3],
                  seq(0.05,max_opac, length.out = length(mybreaks)-1))
  }
  mycols[1] <- missingcol
  data[is.na(data)] <- -.1
  mycolorscheme <- cut(data, mybreaks) %>%
    as.numeric()
  mycolorscheme <- mycols[mycolorscheme]
  levs <- levels(cut(data, mybreaks))
  levs[1] <- "No Data"
  return(list(mycolorscheme, levs, mycols))
}
``` 
   
Let's first use this function to get a map of zip codes where the color of each zip code is related to the positive case rate. For now, we'll divide the case rates into quintiles in order to match the map published by the Furman Center (https://furmancenter.org/thestoop/entry/covid-19-cases-in-new-york-city-a-neighborhood-level-analysis).   
```{r map2, exercise=TRUE, exercise.setup = 'load-data2'}
col_lev <- getShapeColors(round(zcta$Positive/zcta$Total_Pop*1000,1), 
                          5, 
                          "white", 
                          mycolors=viridis(6))

plot(NYC_zips, col = col_lev[[1]], 
     main = "Positive Tests Per 1000 People by Zip Code", 
     cex = 10, lwd=0.01)

legend("topleft", 
       legend = col_lev[[2]], 
       fill = col_lev[[3]], 
       cex = 0.8, 
       title = "Positive Tests per 1000 People")
```   

  
Next, remember the ABC News article that printed a map of zip codes by the proportion of residents with a college degree (article here: https://abcnews.go.com/Health/nyc-stark-contrast-covid-19-infection-rates-based/story?id=69920706)? Let's see if we can reproduce that map. (A note about colors: In the function to get the color scheme -- which is printed above -- there's an input called rdgrbl which allows me to change the color scheme of the map. In the first map, I actually used a built in color scheme called "viridis", which I chose to match the colors in the Furman Center map. This time, I'll use shades of blue to match the ABC News map. The rdgrbl input passes information to the rgb function in R. Basically, you can input three numbers -- 0s or 1s -- corresponding to the colors red, green, and blue, in that order. So blue is represented by the numbers 0,0,1).      
```{r map3, exercise=TRUE, exercise.setup = 'load-data2'}
col_lev <- getShapeColors(zcta$Bach_plus_prop, 
                          5, 
                          "white", 
                          rdgrbl=c(0,0,1),
                          max_opac = .8)

plot(NYC_zips, col = col_lev[[1]], 
     main = "Percent of Population with a College Degree", 
     cex = 10, lwd=0.01)

legend("topleft", 
       legend = col_lev[[2]], 
       fill = col_lev[[3]], 
       cex = 0.8, 
       title = "% Non-White")
```    
  
An open-ended question about trying to visualize these two things on a plot simultaneously: Suppose that we wanted to see how well the two maps above "line up" (without scrolling back and forth between them). Here's one way we might try to do that (although maybe there are better ways.. or maybe this is an ill-advised endeavor altogether): We can plot the case rate quintile as increasingly dark shades of transparent red, and the proportion with a college degree as increasingly dark shades of transparent blue. Finally, we can overlay the maps, one on top of the other. Then any zip code with similar levels of these two variables will appear purple, while zip codes with a high-level of one variable, but not the other, will appear more clearly red or blue. Because the "hot" areas of these two maps do not line up that well, we expect a red/blue map as opposed to a purple one.     

```{r map4, exercise=TRUE, exercise.setup = 'load-data2'}
col_lev1 <- getShapeColors(round(zcta$Positive/zcta$Total_Pop*1000,1), 
                           5, 
                           rdgrbl = c(1,0,0))
col_lev2 <- getShapeColors(zcta$Bach_plus_prop, 
                           5, 
                           rdgrbl = c(0,0,1))

par(mfrow=c(1,3))
plot(NYC_zips, col = col_lev1[[1]], 
     main = "", 
     cex = 10, lwd=0.01)
legend("topleft", 
       legend = col_lev1[[2]], 
       fill = col_lev1[[3]], 
       cex = 0.8, 
       title = "Pos Tests per 1000")

plot(NYC_zips, col = col_lev2[[1]], 
     main = "", 
     cex = 10, lwd=0.01)
legend("topleft", 
       legend = col_lev2[[2]], 
       fill = col_lev2[[3]], 
       cex = 0.8, 
       title = "% College Deg")

plot(NYC_zips, col = col_lev1[[1]], 
     main = "", 
     cex = 10, lwd=0.01)
plot(NYC_zips, col = col_lev2[[1]], 
     main = "", 
     cex = 10, lwd=0.01, add = T)
```   
   
   
We can also look at that plot more up close, and combine it with a scatter plot, which might make the picture a little clearer (note that the red and the blue zones don't really overlap, and most of the points in the scatter plot -- which each represent a zip code -- are either blue or red):  
```{r map5, exercise=TRUE, exercise.setup = 'load-data2'}
col_lev1 <- getShapeColors(round(zcta$Positive/zcta$Total_Pop*1000,1), 
                           5, 
                           rdgrbl = c(1,0,0))
col_lev2 <- getShapeColors(zcta$Bach_plus_prop, 
                           5, 
                           rdgrbl = c(0,0,1))

par(mfrow=c(1,2))
plot(NYC_zips, col = col_lev1[[1]], 
     main = "", 
     cex = 10, lwd=0.01)
plot(NYC_zips, col = col_lev2[[1]], 
     main = "", 
     cex = 10, lwd=0.01, add=T)

plot(zcta$Bach_plus_prop,
     round(zcta$Positive/zcta$Total_Pop*1000,1),
     col = col_lev1[[1]],
     main = "",
     xlab = "Proportion College Degree",
     ylab = "Pos Cases per 1000 Ppl",
     pch = 16)
points(round(zcta$Bach_plus_prop,1),
       round(zcta$Positive/zcta$Total_Pop*1000,1),
       col = col_lev2[[1]],
       pch = 16)
```   
  
If we want to now see what this looks like when the maps DO line up (hot spots in one match up with hot spots in the other, generating more purple), we can use a little trick: Instead of looking at the proportion of people WITH a college degree in each zip code, we can calculate the proportion of people WITHOUT a college degree (exercise: how would you calculate this?). Here is some code to reproduce the map and scatter plot above to make use of this trick. We now see a more purple-colored map and a positive association between the two variables.    
```{r map6, exercise=TRUE, exercise.setup = 'load-data2'}
col_lev1 <- getShapeColors(round(zcta$Positive/zcta$Total_Pop*1000,1), 
                           5, 
                           rdgrbl = c(1,0,0))
col_lev2 <- getShapeColors((100-zcta$Bach_plus_prop), 
                           5, 
                           rdgrbl = c(0,0,1))

par(mfrow=c(1,2))
plot(NYC_zips, col = col_lev1[[1]], 
     main = "", 
     cex = 10, lwd=0.01)
plot(NYC_zips, col = col_lev2[[1]], 
     main = "", 
     cex = 10, lwd=0.01, add=T)

plot((100-zcta$Bach_plus_prop),
     round(zcta$Positive/zcta$Total_Pop*1000,1),
     col = col_lev1[[1]],
     main = "",
     xlab = "Proportion College Degree",
     ylab = "Pos Cases per 1000 Ppl",
     pch = 16)
points((100-zcta$Bach_plus_prop),
       round(zcta$Positive/zcta$Total_Pop*1000,1),
       col = col_lev2[[1]],
       pch = 16)
```   


## Scatter plots   
Let's begin by taking a closer look at some of the scatter plots from earlier sections. For example, we looked at this plot in the "case rates by zip codes" section. The plot shows the positive test rate (positive tests/total tests) on the y-axis and the percent of the population tested (total tests/population).
   
```{r scatter1, ref.label="pop6", exercise=FALSE, exercise.setup = 'load-data'}
```


Recall that we first saw this scatter plot while considering how the *positive case rate* (positive tests/population) is different from a *positive test rate* (positive tests/total tests). In particular, the rank-order of zip code neighborhoods is different depending on which measure we use.

Below we produce two scatterplots side by side, and this time we can add a smoothing curve to highlight the trends.

```{r scatter_smooth1, exercise=TRUE, exercise.setup = 'load-data'}
par(mfrow=c(1,2))
scatter.smooth(zcta$Total/zcta$Total_Pop*100, 
               zcta$zcta_cum.perc_pos,
               span = 0.75,
               xlab = "% of population tested",
               ylab = "positive TEST rate",
               main = "",
               lpars = list(col = "red", lwd = 3))

scatter.smooth(zcta$Total/zcta$Total_Pop*100, 
               zcta$Positive/zcta$Total_Pop*100,
               span = 0.75,
               xlab = "% of population tested",
               ylab = "positive CASE rate",
               main = "",
               lpars = list(col = "red", lwd = 3))

```
  
  
The red line roughly follows the "center" of the data, allowing us to see that increasing and decreasing trend in the test rate more clearly. The case rate does not decrease, although it does start to level off a bit. As it must. The larger our sampling proportion for testing, the closer we get to the true  proportion of positive cases.

However, as you can see clearly from the scatter plots, the highest-ranked zipcodes by test rate are those with testing rates around 3%. In contrast, the zip codes with the highest case rate are those that tested the most.

In neither case can we say that the sample of people tested is representative of the population. First of all, the sickest people are the first ones to get tested. When tests were not widely available, they were also limited to essential workers. This could raise the rate of testing in neighborhoods where those workers live. On the other hand, wealthier people might have  access to testing in private labs, which could raise the rate of testing in wealthier neighborhoods.
  
### More revelations using scatterplots 

We now revisit another scatter plot from the "tables by race and income" section. 

```{r scatter2, ref.label="twoway2", exercise=FALSE}
```

In this plot, we see an overall negative association between case rates and median income. By negative association, we mean that as one variable goes up, the other goes down; in this case, as median income increases, we the case rate tends to decrease. 

However, we also saw from our two-way table that, when we split the zip codes into groups based on proportion white, this relationship between income and case rate varies within groups. Among the zip codes with the lowest proportion of white residents (the darkest blue dots in the plot), there is actually a positive association between median income and case rate. If you look only at the dark blue dots, then it would appear that higher median incomes are actually associated with higher case rates.

To recap: if you drew a line through all the data, it would have a negative slope. However, a line through those dark blue dots would now go from the lower left to upper right (a positive slope). 

Let's actually draw those lines on the plot. The blue and purple lines are smooth curves based only the points with the same color:  

```{r scatter-smooth2, exercise=TRUE, exercise.setup = 'load-data'}
dat <- zcta %>% 
  select(White_Total, Race_Total, Med_Income, Positive, Total_Pop) %>% 
  na.omit() %>%
  mutate(White_Prop = round(White_Total/Race_Total, 3)*100) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 3)) %>%
  mutate(Pos_Case_Rate = Positive/Total_Pop*100)

plot(dat$Med_Income, 
     dat$Pos_Case_Rate, 
     col=dat$White_Prop_Lev+3,
     xlab = "Median Income",
     ylab = "Case Rate",
     pch = 16)

lines(loess.smooth(dat[dat$White_Prop_Lev==1,"Med_Income"],
                   dat[dat$White_Prop_Lev==1,"Pos_Case_Rate"], span=3), col=4, lwd=2)
lines(loess.smooth(dat[dat$White_Prop_Lev==3,"Med_Income"],
                   dat[dat$White_Prop_Lev==3,"Pos_Case_Rate"], span=3), col=6, lwd=2)


legend("topright",
       col = c(4,5,6),
       pch = 16,
       legend = c("Low Prop White",
                  "Med Prop White",
                  "High Prop White"))
```  
  

Now that we've seen how scatter plots can be helpful visualizations when we want to better understand fine-grained relationships between variables, let's use some scatter plots to re-examine the relationship between the racial composition of a given zip code and the case rate. 

We saw previously that there appears to be a negative association between the positive case rate and proportion of the population that identifies as white by zip code. Let us shift our attention to the black and Hispanic population. Reports using data from NYC have shown that black and Hispanic communities are experiencing higher rates of covid infection and mortality; therefore, we expect to see a positive association this time.  

We can visualize this relationship on a scatter plot by representing each zip code as a dot and plotting the positive case rate against the proportion of the population that is black or Hispanic By adding a smoothing line in blue, we can see a positive association between these variables.   

```{r scatter3_bh, exercise=TRUE, exercise.setup = 'load-data'}
scatter.smooth(zcta$NH_Black+zcta$Hisp,
     round(zcta$Positive/zcta$Total_Pop*1000,1), 
     xlab="percent black + Hispanic", 
     ylab="pos cases per 1000 ppl", 
     span=3, lpars=list(lwd=3, col="blue"),
     pch=16)

```
  
Some possible mediators for these racial disparities are income level and, related to income, level of educational attainment. Some have speculated that differences in case rates by race could be "explained" by income and education inequality among different racial groups. If true, this would not mean that varying outcomes are fair, because they are not directly race-based. The difference in health outcomes would still be indirectly race-based. But whereas policy cannot change the racial composition of a neighborhood, it can be directed to improve educational attainment and wealth. Thus, these mediators are levers for change.

To explore whether income and education variables can explain the observed racial disparities, let's first look at the relationship between these variables and case rates. We see that there are negative associations for both median income and college education with case rates.

```{r scatter4, exercise=TRUE, exercise.setup = 'load-data'}
par(mfrow=c(1,2))
scatter.smooth(zcta$Med_Income,
     round(zcta$Positive/zcta$Total_Pop*1000,1), 
     xlab="median income", 
     ylab="pos cases per 1000 ppl", 
     span=3, lpars=list(lwd=3, col="blue"),
     pch=16)

scatter.smooth(zcta$Bach_plus_prop,
     round(zcta$Positive/zcta$Total_Pop*1000,1), 
     xlab="share of college degree", 
     ylab="pos cases per 1000 ppl", 
     span=3, lpars=list(lwd=3, col="blue"),
     pch=16)

```
  
Finally, we can look at the relationship between income and educational outcomes with the racial composition of zip codes. This time, we see that zip codes with a higher proportion of black and Hispanic residents tend to have lower incomes and college attainment. 

```{r scatter5, exercise=TRUE, exercise.setup = 'load-data'}

par(mfrow=c(1,2))
scatter.smooth(
     zcta$NH_Black+zcta$Hisp, 
     zcta$Med_Income,
     xlab="% black or Hispanic", 
     ylab="median income", 
     span=3, lpars=list(lwd=3, col="blue"),
     pch=16)

scatter.smooth(
     zcta$NH_Black+zcta$Hisp, 
     zcta$Bach_plus_prop,
     xlab="% black or Hispanic", 
     ylab="share of college degree", 
     span=3, lpars=list(lwd=3, col="blue"),
     pch=16)


```  
   
This all provides some suggestive evidence that education and income inequality may be mediating some of the racial disparities with regard to case rate. There are more advanced methods of data science (e.g., multiple regression, mediation analysis) to dive deeper into the explanation of racial disparities. But you will need to take a next course to master those!
   
## Exploration workspace  
Now that we've taken a look at some of these variables, this space is left to you for any additional questions you may want to ask. Some code has been copied for you here, which you can run and then alter, in order to address new questions. Questions to think about: What other variables could you look at that might be related to the positive case rate? How are these variables related to each other? How is the number of tests administered related to both the population and positive case rate?  
\newline  
\newline  
### Code to examine the data  
```{r practice1, exercise=TRUE, exercise.setup = 'load-data'}
str(zcta)  
```   
  
### Code to construct a 1-way table   
```{r practice2, exercise=TRUE, exercise.setup = 'load-data'}
inc <- zcta %>% 
  mutate(Med_Inc_Lev = ntile(Med_Income, 3)) %>%
  group_by(Med_Inc_Lev) %>%
  summarise(Cases_per_1000 = round(sum(Positive)/sum(Total_Pop)*1000, 2)) %>%
  na.omit()
inc
```  

### Code to construct a 2-way table of case rates  
```{r practice3, exercise=TRUE, exercise.setup = 'load-data'}
inc_propwhite <- zcta %>% 
  select(White_Total, Race_Total, Med_Income, Positive, Total_Pop) %>% 
  na.omit() %>%
  mutate(White_Prop = round(White_Total/Race_Total, 3)*100) %>%
  mutate(White_Prop_Lev = ntile(White_Prop, 3)) %>%
  mutate(Med_Inc_Lev = ntile(Med_Income, 3)) %>%
  group_by(White_Prop_Lev, Med_Inc_Lev) %>%
  summarise(Cases_per_1000 = round(sum(Positive)/sum(Total_Pop)*1000, 2)) %>%
  spread(White_Prop_Lev, Cases_per_1000, fill=0)

inc_propwhite[,1] <- c("Low Income", "Med Income", "High Income")
colnames(inc_propwhite) <- c("inc/race","Low Prop White", "Med Prop White", "High Prop White")
inc_propwhite
```    

### Code to construct a scatter plot   
```{r practice4, exercise=TRUE, exercise.setup = 'load-data'}
plot(zcta$Med_Income,
     round(zcta$Positive/zcta$Total_Pop*1000,1), 
     xlab="median income", 
     ylab="pos cases per 1000 ppl", 
     pch=16)
```    
  
### Code to construct a map  
```{r practice5, exercise=TRUE, exercise.setup = 'load-data2'}
col_lev <- getShapeColors(round(zcta$Positive/zcta$Total_Pop*1000,1), 
                          5, 
                          "white", 
                          rdgrbl = c(0.3,0.1,0.4),
                          max_opac = 0.9)

plot(NYC_zips, col = col_lev[[1]], 
     main = "Positive Tests Per 1000 People by Zip Code", 
     cex = 10, lwd=0.01)

legend("topleft", 
       legend = col_lev[[2]], 
       fill = col_lev[[3]], 
       cex = 0.8, 
       title = "Positive Tests per 1000 People")
```
   
  
